Visualizing the Geography of Genetic Variants

### 2016-08-16 cabios-template/main.tex
added supplement and clean up main text
Edits to discussion; fixing supp fig/table number

\textbf{Rare variants:} Representing the frequency of rare variants across populations can be difficult as allele frequencies between variants often differ greatly, sometimes by orders of magnitude in a single dataset \citep[e.g.][]{10002015global, nelson2012abundance, tennessen2012evolution}. To address this challenge we re-scale frequencies for rare variants, so that small frequencies become visible. Specifically, we  use a frequency scale that is indicated in a legend below the map and represented by varying color in the pie charts [Figure \ref{fig:screenshot}, Supplementary Figure 1,   Supplementary Table 1].

\textbf{Uncertainty in frequency data:}  One under-appreciated problem with allele frequency maps is that not all data points have equal levels of certainty. To address this issue, we use varying transparency in a population's pie chart: estimated frequencies with higher levels of sampling error (e.g.~those from samples with $n < 30$) are made more transparent, and hence less visible, on the map [Figure \ref{fig:screenshot}, Supplementary Figure 2].

\textbf{Overlapping populations:} In many datasets where populations are sampled densely in geographic space, one problem is that allele frequency plots begin to overlap each other and obscure information. To address this issue, we use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins [Figure \ref{fig:screenshot}, Supplementary Figure 3].  Also, by hovering the mouse cursor over any population, a user can see the population labels and precise frequency information.

\textbf{REST API:} To provide an interface to the population minor allele frequency data, we use a REST API implemented in the python library Flask-RESTful \citep{grinberg2014flask}. The front-end D3.js visualization uses the API to obtain the data, though users can also interface with it directly. Genetic variants can be queried by chromosome position, rsid, or randomly (see supplementary data for examples).

@@

By allowing rapid generation of allele frequency maps, we hope to facilitate the interpretation of variant function and history by practicing geneticists. We emphasize that the overwhelming majority the human genome does not differ  across populations \citep[see a classic treatment in][]{lewontin1972apportionment}. We thus hope that the ability to query random variants from major human population genetic samples will allow students to appreciate the structure of human genetic diversity in an approachable and intuitive form.

>>>>

\textbf{Rare variants:} Representing the frequency of rare variants across populations can be difficult as allele frequencies between variants often differ greatly, sometimes by orders of magnitude in a single dataset \citep[e.g.][]{10002015global, nelson2012abundance, tennessen2012evolution}. To address this challenge we re-scale frequencies for rare variants, so that small frequencies become visible. Specifically, we  use a frequency scale that is indicated in a legend below the map and represented by varying color in the pie charts [Figure \ref{fig:screenshot}, Supplementary Figure S1,   Supplementary Table S1].

\textbf{Uncertainty in frequency data:}  One under-appreciated problem with allele frequency maps is that not all data points have equal levels of certainty. To address this issue, we use varying transparency in a population's pie chart: estimated frequencies with higher levels of sampling error (e.g.~those from samples with $n < 30$) are made more transparent, and hence less visible, on the map [Figure \ref{fig:screenshot}, Supplementary Figure S2].

\textbf{Overlapping populations:} In many datasets where populations are sampled densely in geographic space, one problem is that allele frequency plots begin to overlap each other and obscure information. To address this issue, we use force-directed layouts of the populations such that no two points are overlapping each other, and yet the points will be pulled towards their true origins [Figure \ref{fig:screenshot}, Supplementary Figure S3].  Also, by hovering the mouse cursor over any population, a user can see the population labels and precise frequency information.

\textbf{REST API:} To provide an interface to the population minor allele frequency data, we use a REST API implemented in the python library Flask-RESTful \citep{grinberg2014flask}. The front-end D3.js visualization uses the API to obtain the data, though users can also interface with it directly. Genetic variants can be queried by chromosome position, rsid, or randomly (see Supplementary appendix for examples).

@@

By allowing rapid generation of allele frequency maps, we hope to facilitate the interpretation of variant function and history by practicing geneticists. Also, for students of human diversity, it is often difficult to conceptualize classic statements regarding how most variation in humans is shared among populations \citep{lewontin1972apportionment} and how the fixation index $F_{ST}$ is relatively low globally \citep[10-15\%][]{10002015global}.  We hope that the ability to query random variants from major human population genetic samples will allow students to appreciate the structure of human genetic diversity in an approachable and intuitive form.  One caveat of these maps is that one must interpret the geographic positioning of the plotted populations with care (see Supplementary Information for expanded discussion of this point).
